---
title: "Jordan et al. (2022) R Analysis Notebook"
output: html_notebook
---

```{r Required libraries}
library(tidyverse)
library(agricolae)
library(ggpubr)
library(cowplot)
```
<br>

```{r Load in data}
counts_df <- read_csv("competition_data.csv")
counts_df$paired_ID <- as.character(counts_df$paired_ID)
strain_MICs_df <- read_csv("AMP_strains_MICs.csv") # MIC data from Card et al. 2019
```
<br>

```{r Data wrangling}
fitness_df <- counts_df %>%
  filter(!comp1_d3 < 1) %>% # Removes one of the KJC128 replicates because we did not observe it on the competition plates by day 3
  mutate(malth_comp1 = log(comp1_d3 * (100^3) / comp1_d0)) %>%
  mutate(malth_comp2 = log(comp2_d3 * (100^3) / comp2_d0)) %>% 
  mutate(relative_fitness = malth_comp1 / malth_comp2) %>% 
  select(!c(antibiotic, competitor_2, comp1_d0:comp2_d3, malth_comp1, malth_comp2))

CompareFitness <- function(dat) {
  fitness_vec <- c()
  
  for (i in unique(dat$block)) {
    block_frame <- dat %>% 
      filter(block == i)
    
    for (j in unique(block_frame$paired_ID)) {
      paired_frame <- block_frame %>% 
        filter(paired_ID == j)
      
      true_fitness <- paired_frame[1, 6] / paired_frame[2, 6]

      fitness_vec <- bind_rows(fitness_vec, true_fitness)
    }
  }
  return(fitness_vec)
}

# Normalizes relative fitness based upon common competitor values
normalized_fitness_col <- CompareFitness(fitness_df) %>% 
  drop_na %>% 
  transmute(ln_relative_fitness = log(relative_fitness)) # Log(e) transform the relative fitness values

fitness_df <- fitness_df %>%
  drop_na %>%
  bind_cols(., normalized_fitness_col) %>% 
  select(!paired_ID:relative_fitness) %>% 
  rename(strain = competitor_1)

fitness_df$background <- as.factor(fitness_df$background)
fitness_df$strain <- as.factor(fitness_df$strain)

# Creates new data frame with average fitness and number of replicates for each strain
JoinedDF <- function(dat) {
  mean_sd_df <- dat %>%
    group_by(strain) %>% 
    summarize(average = mean(ln_relative_fitness))
    
  replicates_df <- dat %>% 
    group_by(strain) %>% 
    count()
    
  joined_df <- left_join(mean_sd_df, replicates_df, by = "strain")
}

joined_df <- JoinedDF(fitness_df)

# Creates a data frame with summary statistics  
strain_vec <- c("KJC108", "KJC109", "KJC110", "KJC111", "KJC112", "KJC113", "KJC114", "KJC115", "KJC120", "KJC121", "KJC122", "KJC123", "KJC128", "KJC129", "KJC130", "KJC131")

SummaryStats <- function(dat, strains) {
  strain_frame <- dat %>% 
    filter(strain == strains)
  
  if (strain_frame$n == 5) {
    mutated_frame <- strain_frame %>%
      mutate(lower_CI = average - (2.776 * (sqrt(0.00441) / sqrt(n)))) %>%  # Error bars calculated using the t-distribution, 4 d.f., and pooled SD estimated from ANOVA.
      mutate(upper_CI = average + (2.776 * (sqrt(0.00441) / sqrt(n))))
  } else {
    mutated_frame <- strain_frame %>%
      mutate(lower_CI = average - (3.182 * (sqrt(0.00441) / sqrt(n)))) %>%  # Error bars calculated using the t-distribution, 3 d.f., and pooled SD estimated from ANOVA.
      mutate(upper_CI = average + (3.182 * (sqrt(0.00441) / sqrt(n))))
  }
  return(mutated_frame)
}

summary_stats_df <- map_df(strain_vec, ~SummaryStats(joined_df, .x))
```
<br>

**Analyses**

One-sample t-test of average fitness costs of resistant lines
```{r Average fitness cost}
t.test(summary_stats_df$average, mu = 0, alternative = "less")
```
<br>

ANOVA tests of fitness differences among strains
```{r Test for significant variation among lines}
FitnessAnova <- function(dat, tukey = FALSE) {
    aov <- aov(ln_relative_fitness ~ strain, data = dat)
    
    if (tukey == FALSE) {
      return(summary(aov))
    }
    else {
      return(aov)
      }
    }

FitnessAnova(fitness_df)
```
<br>

More data wrangling - Merge summary stats with MIC values from Card et al. 2019
```{r Yet more data wrangling}
combined_df <- strain_MICs_df %>%
  left_join(summary_stats_df, strain_MICs_df, by = "strain") %>%
  mutate(MIC_parent = log2(MIC_parent)) %>% 
  mutate(MIC_daughter = log2(MIC_daughter)) %>% 
  mutate(fold_change = (MIC_daughter - MIC_parent))
```
<br>

Correlations between strain fitness and MIC
```{r Correlation between fitness and MIC}
cor.test(combined_df$average, combined_df$MIC_daughter, alternative = "two.sided")
```
<br>

Correlations between strain fitness and level of resistance conferred by mutation
```{r Correlation between fold-change in resistance and MIC}
cor.test(combined_df$average, combined_df$fold_change, alternative = "two.sided")
```
<br>

ANOVA to test for main effect of genetic background on average fitness costs
```{r Test for effect of background on fitness costs}
aov(average ~ background, data = combined_df) %>% summary()
```
<br>

ANOVA to test for possible interaction between genetic background and MIC on average fitness costs
```{r Test for interaction effects}
aov(average ~ background * MIC_daughter, data = combined_df) %>% summary()
```
<br>

Compare resistant lines with single mutations against lines with multiple mutations (i.e., hitchhiking hypothesis)
```{r Test for genetic-hitchhiking effects}
avg_fitness_mult_df <- combined_df %>% 
  filter(strain %in% c("KJC109", "KJC114"))

avg_fitness_single_df <- anti_join(combined_df, avg_fitness_mult_df)

mean(avg_fitness_single_df$average)
mean(avg_fitness_mult_df$average)

t.test(avg_fitness_single_df$average, avg_fitness_mult_df$average, alternative = "greater")
```
<br>


Tukey test
```{r Post-hoc test}
tukey_test <- FitnessAnova(fitness_df, tukey = TRUE) %>% 
  HSD.test(., trt = "strain")

tukey_groups <- data.frame(group = c("ef", "bcd", "cde", "ef", "ab", "ab", "ab", "def", "ef", "ef", "a", "f", "ef", "abc", "f", "ef"))
combined_df <- bind_cols(combined_df, tukey_groups)
```
<br>


**Visualizations of the data**

```{r Correlation plots}
fitness_versus_MIC_plot <- combined_df %>% 
  ggplot(aes(x = MIC_daughter, y = average)) +
    geom_point(size = 2) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "black") +
    stat_cor(method = "pearson", cor.coef.name = "r", digits = 4, label.x = 0.9, label.y = -0.08, geom = "text", size = 5) +
    scale_x_continuous(limits = c(0, 2), breaks = c(0, 1, 2)) +
    labs(y = "Log"[e]~"relative fitness", x = "Log"[2]~"MIC") +
    theme_cowplot()

fitness_versus_foldMIC_plot <- combined_df %>% 
  ggplot(aes(x = fold_change, y = average)) +
    geom_point(size = 2) +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "black", size = 1) +
    stat_cor(method = "pearson", cor.coef.name = "r", digits = 4, label.x = 1.4, label.y = -0.08, geom = "text", size = 5) +
    scale_x_continuous(limits = c(0, 3), breaks = c(0, 1, 2, 3)) +
    labs(y = "Log"[e]~"relative fitness", x = "Log"[2]~"fold-increase in MIC") +
    theme_cowplot() +
    theme(axis.title.y = element_text(color = "white"),
        axis.text.y = element_text(color = "white"),
        axis.ticks.y = element_blank())

corr_plot <- plot_grid(fitness_versus_MIC_plot, fitness_versus_foldMIC_plot,
                       labels = "AUTO")

corr_plot

# ggsave("corr_plot.pdf", corr_plot, path = "Figures", device = "pdf", width = 10, height = 6, units = "in")
# ggsave("corr_plot.png", corr_plot, path = "Figures", device = "png", width = 10, height = 6, units = "in")
```
<br>

```{r Average fitness plot}
avg_fitness_plot <- combined_df %>% 
  ggplot(aes(x = strain, y = average)) +
    geom_point(aes(x = reorder(strain, average) , y = average), size = 3) +
    geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = 0.3) +
    geom_text(aes(label = group), vjust = -7) +
    scale_y_continuous(limits = c(-0.5, 0.2), breaks = c(-0.5, -0.4, -0.3, -0.2, -0.1, 0.0, 0.1, 0.2)) +
    scale_x_discrete(labels = c("Ara\u20135-2", "Ara\u20135-1", "Ara+5-1", "Ara+4-1", "Ara+5-3", "Ancestor-2", "Ancestor-3", "Ara\u20136-1", "Ancestor-4",
                                "Ara+4-2", "Ara+4-3", "Ara\u20136-3", "Ancestor-1", "Ara+5-2", "Ara\u20135-3", "Ara\u20136-2")) +
    labs(y = expression("Log"[e]~"relative fitness")) +
    theme_cowplot() +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

avg_fitness_plot

# ggsave("avg_fitness_plot.pdf", avg_fitness_plot, path = "Figures", device = "pdf", width = 10, height = 6, units = "in")
# ggsave("avg_fitness_plot.png", avg_fitness_plot, path = "Figures", device = "png", width = 10, height = 6, units = "in")
```